D3064R0: OOTA Execution is Provably Vacuous

### 2023-11-29 ootavacuous.tex
Apply David Goldblatt feedback
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
The Indefatigible TBD
}
\date{November 27, 2023 (Pre-Tokyo)}
\maketitle{}

@@

Listing~\ref{lst:OOTA Cycle}
shows an OOTA cycle, where the \co{\{\{\{} on line~5 begins a set of
threads, the \co{|||} on line~8 separates a pair of threads, and
the \co{\}\}\}} on line~9 ends a set of threads.
Lines~6 and~7 do a relaxed copy from \co{x} to \co{y}, and
lines~8 and~9 to a relaxed copy from \co{y} to \co{x}.
Even though both \co{x} and \co{y} are initialized to zero on
lines~2 and~3, respectively, according to the mathematical core
of the C++ memory model, the only constraint on the final values
of \co{x} and \co{y} is that they be equal.
Counterintuitivily, their final values could be any member of their type.

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work, and by the laws of physics, which have
not.

@@

Because C++ volatile atomics constitute observable behavior,\footnote{
	Working Draft, Standard for Programming Language C++
	\co{[intro.abstract]}.}
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any C++ volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

@@

@@

>>>>

\author{
Paul E.~McKenney\\\email{paulmck@kernel.org} \and
Michael Wong\\\email{fraggamuffin@gmail.com} \and
Maged Michael\\\email{maged.michael@gmail.com} \and
The Indefatigible TBD
}
\date{November 27, 2023 (Pre-Tokyo)}
\maketitle{}

@@

Listing~\ref{lst:OOTA Cycle}
shows an OOTA cycle, where the \co{\{\{\{} on line~5 begins a set of
threads, the \co{|||} on line~8 separates a pair of threads, and
the \co{\}\}\}} on line~9 ends a set of threads.\footnote{
	This is pseudocode in the format of CPPMEM,
	which does not permit writes from non-local variables.}
Lines~6 and~7 do a relaxed copy from \co{x} to \co{y}, and
lines~8 and~9 to a relaxed copy from \co{y} to \co{x}.
Even though both \co{x} and \co{y} are initialized to zero on
lines~2 and~3, respectively, according to the mathematical core
of the C++ memory model, the only constraint on the final values
of \co{x} and \co{y} is that they be equal.
Counterintuitivily, their final values could be any member of their type.

\begin{listing}[tbp]
\begin{verbatim}
 1 int main() {
 2   atomic_int x = 0;
 3   atomic_int y = 0;
 4   volatile int r1;
 5   volatile int r2;
 6   {{{ { r1 = x.load(memory_order_relaxed);
 7         if (r1 == 1)
 8           y.store(1, memory_order_relaxed);
 9         else
10           y.store(r1, memory_order_relaxed); }
11   ||| { r2=y.load(memory_order_relaxed);
12         x.store(r2,memory_order_relaxed); }  }}}
13   }}};
14   return 0;
15 }
\end{verbatim}
\caption{OOTA Cycle Based on Conditionals}
\label{lst:OOTA Cycle Based on Conditionals}
\end{listing}

A similar effect can be achieved using conditional statements, as
shown in Listing~\ref{lst:OOTA Cycle Based on Conditionals}.

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work, and by the laws of physics, which have
not.
Additional constraints are imposed by hardware architecture and
design, which have been partially accounted for in the standard,
one limitation being that accurate and executable formal descriptions
of hardware memory models did not appear until after the standard
was released.

@@

Because C++ volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2022N4910},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any C++ volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

\subsection{Hardware Architecture and Design}
\label{sec:Hardware Architecture and Design}

Heavily used CPU families now have accurate and executable formal
memory models, but even more important, there is much more clarity
on exactly what hardware will and will not do in the course of
speculative execution.

\begin{listing}[tbp]
\begin{verbatim}
 1 int foo(int i) {
 2   atomic_int x = 0;
 3   int y[3];
 4   volatile int r1;
 5   {{{ { y[i].store(42, memory_order_relaxed);
 6         x.store(1, memory_order_relaxed); }
 7   ||| r1 = x.load(memory_order_relaxed);
 8   }}};
 9   return 0;
10 }
\end{verbatim}
\caption{Speculated Store and Non-Speculated Load}
\label{lst:Speculated Store and Non-Speculated Load}
\end{listing}

For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{a[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
	nor cares.}
Then line~7 can cause the final value of \co{r1} to be 42, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

Therefore, non-speculative loads are prohibited from loading values from
a speculative stores.

In addition, in order to avoid hardware-based OOTA, a value stored
speculatively will not be exposed to loads from threads other than the
one executing that store.

@@

\subsection{Illustrative Examples}
\label{sec:Illustrative Examples}

This section works through OOTA examples, showing the effects of
the constraints called out in this paper.

\subsubsection{OOTA Cycle}
\label{sec:OOTA Cycle}

Listing~\ref{lst:OOTA Cycle}
shows the canonical example of an OOTA cycle.
Because values from speculative stores cannot be visible to other threads,
line~7 cannot be committed until after line~6 completes.
Similarly, line~9 cannot be committed until after line~8 completes.

Because values stored take non-zero time to propagate from one thread
to another, no cycle can form, and thus no OOTA cycle can form.

\subsubsection{OOTA Cycle Based on Conditionals}
\label{sec:OOTA Cycle Based on Conditionals}

Listing~\ref{lst:OOTA Cycle Based on Conditionals}
shows an if-condition-based example of an OOTA cycle.
Because values from speculative stores cannot be visible to other threads,
lines~8 and~10 cannot be committed until after the load in line~6 completes
and the conditional on line~7 is evaluated non-speculatively.
Similarly, line~12 cannot be committed until after line~11 completes.

Because values stored take non-zero time to propagate from one thread
to another, no cycle can form, and thus no OOTA cycle can form.

Note that if both legs of the \co{if} statement spanning lines~7-10,
the compiler would be within its rights to hoist the store out of
that \co{if} statement.
However, in this case, the two stores on lines~8 and~10 are storing
distinct values, preventing the compiler from hoisting.

Note also that some architectures provide conditional-move instructions,
which allows line~7 to avoid emitting a conditional branch.
However, such architectures would still treat a store of the
target of the conditional-move instructions as speculative until
such time as the conditional moves (and any loads that they depend
on) complete.

@@

\section{Acknowledgments}
\label{sec:Acknowledgments}

We are grateful to David Goldblatt for his careful review of an early
draft of this paper.
Nonetheless, all errors and omissions in this paper are the sole property
of the authors.
